MSF_Groupwork-NBA_shotlog

### 2016-10-16 MSFGroupwork-NBAHotHand.Rmd
final tidying up

The NULL Hypothesis: Analysis of means for home and away wins does not show any significant difference in the means.

@@

```{r, echo=FALSE}
graph.acc <- ggplot(acc, aes(x = seconds.left, y = accuracy)) + geom_point() + geom_smooth()

@@

```{r, echo=FALSE}

@@

## Shooting Percentage Analysis (James)

@@

```{r, echo=FALSE}
graph01 <- ggplot(special2) + geom_boxplot(aes(x = player_name, y = FGA, colour = W))
graph01 <- graph01 + labs(title = "Differences in FGA for games lost/won of the three special players") + xlab("Player name") + ylab("FGA")
graph02 <- ggplot(special2) + geom_boxplot(aes(x = player_name, y = FG.percentage, colour = W)) 
graph02 <- graph02 + labs(title = "Differences in FG% for games lost/won of the three special players") + xlab("Player name") + ylab("FG%")
grid.arrange(graph01, graph02, ncol = 2)
```

@@

###Plot Result
  The explanation from the plot above is summerised in the following table:

** Table 13**: Relationship between the defender distance and accuracy

Defender Distance (range in feet) | accuracy (range in %) 
- | -
0 - 2 | 42 - 50
2 - 4 | 44 - 49
4 - 6 | 41 - 46
6 - 8 | 37 - 42
8 - 10 | 37 - 54
10 - 16 | 33 - 60
16+ | 0 - 100

@@

**Table 14**: Relationship between the shooter distance ranges and accuracy

  Shooter Distance (in feet) | Accuracy (range in %)
  - | -
  0 - 2 | 62 - 68
  2 - 4 | 61 - 67
  4 - 6 | 44 - 58
  6 - 8 | 38 - 51
  8 - 10 | 35 - 42
  10 - 12 | 34 - 42
  12 - 14 | 37 - 47
  14 - 16 | 33 - 51
  16 - 18 | 37 - 48
  18 - 24 | 33 - 43
  24 - 28 | 18 - 38
  28+ | 0 - 50

@@

##Hot Hand Theory
###Introduction
  The analysis aims to investigate whether the effect of the 'hot hand theory' is valid based on the given dataset. The 'hothand theory' states that if the shooter makes his first shot, the probability of making his next shot is higher.

  The specific analysis aims to show the effect of the 'hot hand theory' on NBA players. The analysis is aimed to establish whether there are  the hypothesis that a shooter with a hot hand will have shots that are dependent of one another. If the shooter makes his first shot, the probability of making his next shot is higher.

@@

```{r}
#Showing distribution for each player's streak lengths
gg1 <- ggplot(data = player1_df, aes(x = streaks)) + geom_bar(fill = 'brown') + labs(title = 'Kyrie Irnving')
gg2 <- ggplot(data = player2_df, aes(x = streaks)) + geom_bar(fill = 'red') + labs(title = 'Mo Williams')
gg3 <- ggplot(data = player3_df, aes(x = streaks)) + geom_bar(fill = 'blue') + labs(title = 'Klay Thompson')
grid.arrange(gg1, gg2, gg3, ncol = 3)
```

@@

```{r}
#Create a boxplot for each player's streak lengths 
attach(mtcars)
par(mfrow=c(1, 3))
boxplot(player1_df$streaks, main = 'Kyrie Irnving')
boxplot(player2_df$streaks, main = 'Mo Williams')
boxplot(player3_df$streaks, main = 'Klay Thompson')
```

@@

```{r}
#Comparing barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar(fill = 'blue') + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar(fill = 'red') + labs(title = 'Kyrie Irving')
grid.arrange(g1, g2, ncol = 2)
```

####Plot Result
  After running many times the sample, there are several times when the distributions of sample and real shooter are quite similar. As a consequence, it can be concluded that Mo Williams does not have a hot hand.

### Mo Williams
```{r, include=FALSE}
###PLAYER 2 VS SAMPLE PLAYER###
#Simulation - creating a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter2 
set.seed(572)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[2]]), replace = TRUE, 
                        prob = c(shooting_percentage[[2]], 1 - shooting_percentage[[2]]))
simul_shooter

@@

```{r}
#Comparing barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar(fill = 'blue') + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar(fill = 'green') + labs(title = 'Mo Williams')
grid.arrange(g1, g2, ncol = 2)
```

@@

  After running many times the sample, there are several times when the distributions of sample and real shooter are quite similar. As a consequence, it can be concluded that Mo Williams does not have a hot hand.

###Klay Thompson
```{r, include=FALSE}
###PLAYER 3 VS SAMPLE PLAYER###
#Simulation - creating a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter3 
set.seed(57)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = length(streak_per_player[[3]]), replace = TRUE, 
                        prob = c(shooting_percentage[[3]], 1 - shooting_percentage[[3]]))
simul_shooter

@@

```{r}
#Comparing barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar(fill = 'blue') + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar(fill = 'brown') + labs(title = 'Klay Thompson')
grid.arrange(g1, g2, ncol = 2)
```

@@

  Klay Thompson is an exception making 13 shots in a row. There is no such distribution in the sample, so it could be assumed that he is likely to have a hot hand. Further statistical analysis - logistic regression - should be done in order to reach a more accurate conclusion.

@@

  As the analysis shown, there are some extremely high shot streaks achieved from the players. Kyrie Irnving achieved six shots in a row, Mo Williams 4 shots in a row - twice, while Klay Thompson 13 shots in a row. However, to assume that a player has a hot hand, it is required to prove that his continuous shots are dependent of one another. As the vectors of streaks do not contain continuous variables, a linear regression cannot be used. Creating a sample independent shooter with same number of attempts and same shooting percentage as the real's one, is the closest approach to the specific question. In the first two cases, there are several times when the distribution of streaks of the sample independent player is similar to the real shooter's distribution. As the sample shooter's distribution visualises independent shots, it can be concluded that Kyrie Irnving's and Mo Williams' continuous shots are independent of one another and as a result they have not a hot hand. On the other hand, Klay Thompson is likely to have a hot hand, as his shots length distribution diverges significantly from sample's distribution.

@@

  Since the dataset is reffered to only one season and the analysis is not able to eliminate the hot hand theory, it would be an interesting topic to be investigated in future seasons.

```{r, include=FALSE}
#Regression Analysis

df1$GAME_CLOCK <- as.numeric(df1$GAME_CLOCK)

df1$AWAY_TEAM <- as.character(df1$AWAY_TEAM)

linreg_fgm <- lm(FGM ~ LOCATION + W + FINAL_MARGIN + SHOT_NUMBER + PERIOD + GAME_CLOCK + SHOT_CLOCK+ DRIBBLES + TOUCH_TIME + SHOT_DIST + PTS_TYPE + CLOSEST_DEFENDER_PLAYER_ID + CLOSE_DEF_DIST + player_id + WINNER, df1)

summary(linreg_fgm)

linreg_fgm2 <- lm(FGM ~ FINAL_MARGIN + SHOT_NUMBER + PERIOD + SHOT_CLOCK + DRIBBLES + TOUCH_TIME + SHOT_DIST + PTS_TYPE + CLOSE_DEF_DIST + player_id, df1 )

summary(linreg_fgm2)

@@

  Ahart (1973). In Kendall, P. and Hollon, S. (1979). Cognitive-behavioral interventions. New York: Academic Press.
  Economist: As sweet as ever (2015) Available at: http://www.economist.com/blogs/gametheory/2015/06/home-advantage-basketball (Accessed: 16 October 2016).
  Erculj, F. and Supej, M. (2009). Impact of Fatigue on the Position of the Release Arm and Shoulder Girdle over a Longer Shooting Distance for an Elite Basketball Player. Journal of Strength and Conditioning Research, 23(3), pp.1029-1036.
  Mascret, N., Ibáñez-Gijón, J., Bréjard, V., Buekers, M., Casanova, R., Marqueste, T., Montagne, G., Rao, G., Roux, Y. and Cury, F. (2016). The Influence of the ‘Trier Social Stress Test’ on Free Throw Performance in Basketball: An Interdisciplinary Study. PLOS ONE, 11(6), p.e0157215.

>>>>

The NULL Hypothesis: Analysis of means for home and away wins does not show any significant difference

@@

```{r, echo=FALSE, warning=FALSE}
graph.acc <- ggplot(acc, aes(x = seconds.left, y = accuracy)) + geom_point() + geom_smooth()

@@

```{r, echo=FALSE,warning=FALSE}

@@

## Game result impact by field goal attempt

@@

```{r, echo=FALSE}
graph01 <- ggplot(special2) + geom_boxplot(aes(x = player_name, y = FGA, colour = W))
graph01 <- graph01 + labs(title = "Differences in FGA for games lost/won /n of the three special players") + xlab("Player name") + ylab("FGA")
graph02 <- ggplot(special2) + geom_boxplot(aes(x = player_name, y = FG.percentage, colour = W)) 
graph02 <- graph02 + labs(title = "Differences in FG% for games lost/won /n of the three special players") + xlab("Player name") + ylab("FG%")
grid.arrange(graph01, graph02, ncol = 2)
```

@@

###Plot Result
  The explanation from the plot above are as follows:   
  1. Defender distance range from 0 to 2 feet, the accuracy varied between 42% to 50%  
  2. Defender distance range from 2 to 4 feet, the accuracy rate for this distance range from 44% to 49%  
  3. Defender distance range from 4 to 6 feet, the accuracy rate for this distance range from 41% to 46%
  4. Defender distance range from 6 to 8 feet, the accuracy rate for this distance range from 37% to 42%
  5. Defender distance range from 8 to 10 feet, the accuracy rate for this distance range from 37% to 54%
  6. Defender distance more than 10 feet up until 16 feet, the accuracy varies greater than previous distance range, starting from 33% to 60%.  
  7. Distance greater than 16 feet, the data varies between 0% (no shots scored from that distance) to 100% (certain successfull shot from that distance)

@@

**Table 13**: The relationship between the shooter distance ranges and accuracy

  Distance (in feet) | Accuracy (range in %)
  - | -
  0 - 2 | 62 - 68
  2 - 4 | 61 - 67
  4 - 6 | 44 - 58
  6 - 8 | 38 - 51
  8 - 10 | 35 - 42
  10 - 12 | 34 - 42
  12 - 14 | 37 - 47
  14 - 16 | 33 - 51
  16 - 18 | 37 - 48
  18 - 24 | 33 - 43
  24 - 28 | 18 - 38
  28+ | 0 - 50

@@

##Hot Hand Theory
###Introduction

  The analysis aims to investigate whether the effect of the 'hot hand theory' is valid based on the given dataset. The 'hothand theory' states that if the shooter makes his first shot, the probability of making his next shot is higher. (Gilovich, T., Vallone, R. and Tversky, A. (1985))

@@

```{r, echo=FALSE}
#Showing distribution for each player's streak lengths
gg1 <- ggplot(data = player1_df, aes(x = streaks)) + geom_bar(fill = 'brown') + labs(title = 'Kyrie Irnving')
gg2 <- ggplot(data = player2_df, aes(x = streaks)) + geom_bar(fill = 'red') + labs(title = 'Mo Williams')
gg3 <- ggplot(data = player3_df, aes(x = streaks)) + geom_bar(fill = 'blue') + labs(title = 'Klay Thompson')
grid.arrange(gg1, gg2, gg3, ncol = 3)
```

@@

```{r, include=FALSE}
#Create a boxplot for each player's streak lengths 
attach(mtcars)
par(mfrow=c(1, 3))
boxplot(player1_df$streaks, main = 'Kyrie Irnving')
boxplot(player2_df$streaks, main = 'Mo Williams')
boxplot(player3_df$streaks, main = 'Klay Thompson')
```

@@

```{r, echo=FALSE}
#Comparing barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar(fill = 'blue') + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar(fill = 'red') + labs(title = 'Kyrie Irving')
grid.arrange(g1, g2, ncol = 2)
```

####Plot Result
  After producing a sample of shots for a player, the result shows distributions of sample and real shooter are quite similar. As a consequence, it can be concluded that Kyrie Irving does not have a hot hand.

### Mo Williams
```{r, include=FALSE}
###PLAYER 2 VS SAMPLE PLAYER###
#Simulation - creating a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter2 
set.seed(572)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = sum(best_players_df[2,4], best_players_df[2,5], best_players_df[2,6]), replace = TRUE, 
                        prob = c(shooting_percentage[[2]], 1 - shooting_percentage[[2]]))
simul_shooter

@@

```{r, echo=FALSE}
#Comparing barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar(fill = 'blue') + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar(fill = 'green') + labs(title = 'Mo Williams')
grid.arrange(g1, g2, ncol = 2)
```

@@

  After producing a sample of shots for a player, the result shows distributions of sample and real shooter are quite similar. As a consequence, it can be concluded that Mo Williams does not have a hot hand.

###Klay Thompson
```{r, include=FALSE}
###PLAYER 3 VS SAMPLE PLAYER###
#Simulation - creating a vector of shots('made', 'missed') for a sample shooter, setting as probability the shooting percentage of the real shooter3 
set.seed(57)
shot_results <- c('made', 'missed')
simul_shooter <- sample(shot_results, size = sum(best_players_df[3,4], best_players_df[3,5], best_players_df[3,6]), replace = TRUE, 
                        prob = c(shooting_percentage[[3]], 1 - shooting_percentage[[3]]))
simul_shooter

@@

```{r, echo=FALSE}
#Comparing barplots of sample shooter and real shooter1
#If real shooter's streaks diverge significantly from an independent shooter's streaks(sample shooter's streaks), we can conclude tha the real shooter has a 'hot hand'. 
g1 <- ggplot(data = simulshooter_df, aes(x = streaks )) + geom_bar(fill = 'blue') + labs(title = 'sample shooter')
g2 <- ggplot(data = realshooter_df, aes(x = streaks)) + geom_bar(fill = 'brown') + labs(title = 'Klay Thompson')
grid.arrange(g1, g2, ncol = 2)
```

@@

  Klay Thompson is an exception making 13 shots in a row. There is no such distribution in the sample, so it could be assumed that he is likely to have a hot hand. However, further analysis of Klay Thompson's (and other basketball players) shooting streaks in other games is required to see if such a rare occurence is repeated(this is out of scope for this analysis).

@@

  The analysis demonstrates that the streaks generated by the players (K. Irving & M. Williams) are similar to randomly generated streaks, which use these players shooting probabilities. As a result there is no evidence to suggest that there is a hot hand phenomenon present here.

  In the case of K. Thompson, his streak is seem as somewhat anomaly and further research is required to see if any similar results have been achieved by the player.

@@

  Since the dataset is reffered to only one season and the analysis is not able to eliminate the hot hand theory, it would be an interesting topic to be investigated by using data from more than one season.

@@

  Ahart (1973). In Kendall, P. and Hollon, S. (1979). Cognitive-behavioral interventions. New York: Academic Press.
  Economist: As sweet as ever (2015) Available at: http://www.economist.com/blogs/gametheory/2015/06/home-advantage-basketball (Accessed: 16 October 2016).
  Erculj, F. and Supej, M. (2009). Impact of Fatigue on the Position of the Release Arm and Shoulder Girdle over a Longer Shooting Distance for an Elite Basketball Player. Journal of Strength and Conditioning Research, 23(3), pp.1029-1036.
  Gilovich, T., Vallone, R. and Tversky, A. (1985) ‘The hot hand in basketball: On the misperception of random sequences’, Cognitive Psychology, 17(3), pp. 295–314. doi: 10.1016/0010-0285(85)90010-6.
  Mascret, N., Ibáñez-Gijón, J., Bréjard, V., Buekers, M., Casanova, R., Marqueste, T., Montagne, G., Rao, G., Roux, Y. and Cury, F. (2016). The Influence of the ‘Trier Social Stress Test’ on Free Throw Performance in Basketball: An Interdisciplinary Study. PLOS ONE, 11(6), p.e0157215.
